R Course Syllabus

### 2023-04-17 Draft Syllabus.Rmd
continuing work on lesson 5

#### Conditionals and their Use

@@

```{r l5_1_1}
weather <- sample(c("rainy", "sunny"), 1)
if (weather == "rainy") {
  umbrella <- "did"
} else {
  umbrella <- "did not"
}

paste("The weather is", weather, "so I", umbrella, "bring my umbrella.")
```

The sentence changes depending on the weather.

Here is an example with more applicability to data analysis. Suppose that you are conducting a study where some students are randomly assigned to attend private school. You want to compare the test scores of students assigned to private school and the test scores of students who remained in a public school.

@@

>>>>

#### **Conditionals and their Use**

@@

```{r l5_1_1}
# Weather vector/variable, storing the day's weather
weather <- sample(c("rainy", "sunny"), 1)
# If/then conditional
if (weather == "rainy") {
  umbrella <- "brought my umbrella"
} else {
  umbrella <- "did not bring my umbrella"
}
# Displaying the result
paste("The weather is", weather, "so I", umbrella)
```

First, we created a variable to hold the weather. The code above just randomly chooses the weather as either rainy or sunny. Then, we created the conditional. `if` tells R that we want an if conditional. We enclose the condition in parentheses `(weather == rainy)`. The curly braces `{}` tell R what to do in the event that it is rainy. Then `else {}` tells R what to do in the event that the weather is not rainy. The final output changes depending on the weather. 
You are not limited to just two options. There is also `else if`.

```{r}
# `rm()` function removes an object from an environment
rm(umbrella)
weather <- "cloudy"
# If/then conditional
if (weather == "rainy") {
  umbrella <- "brought my umbrella"
} else if (weather == "cloudy") {
  umbrella <- "brought my umbrella"
} else {
  umbrella <- "did not bring my umbrella"
}
# Displaying the result
paste("The weather is", weather, "so I", umbrella)
```

You could stack multiple `else if` statements as your `if` conditional becomes more and more complex. There is also a vectorized version of if/else `ifelse()`. Use this function when you'd like to operate on a set of values at the same time. `ifelse()` is particularly useful when manipulating data.

Here is an example with more applicability to data analysis. Suppose that you are conducting a study where some students are randomly assigned to attend private school. You want to compare the test scores of students assigned to private school and the test scores of students who remained in a public school. When a student is assigned to a private school the treatment variable is set to 1 and when they are not it is set to 0.

@@

```{r l5_1_2}
# Sampling a treatment variable and creating testscores
study_data <- list(treatment = sample(c(0, 1), 1000, replace = T), score = round(runif(1000, 0, 100)))
study_data <- as.data.frame(study_data)

# table showing the distribution of treated and untreated cases
table(study_data$treatment)

# Summary of student schore
summary(study_data$score)

# Graph of scores
study_data %>%
  mutate(treatment = factor(treatment, levels = c(0, 1), labels = c("No", "Yes"))) %>%
  group_by(treatment) %>%
  ggplot(aes(x = score, fill = treatment)) +
  geom_bar(alpha = .5) +
  facet_grid(. ~ treatment)
```

Suppose that there was another piece of your study. All students who did not receive the treatment were instead provided with free metro cards. Some of the private school students got free metro cards, but they got them through other sources. We can represent this in the data using `ifelse()` and `mutate()`.

```{r l5_1_3}
private_students <- nrow(study_data[study_data$treatment == 1,])
study_data <-
  study_data %>%
    # In English: If a student was not treated then they received a metro card, otherwise they may have had a metro card and generate those values randomly
    mutate(metro_card = ifelse(treatment == 0, 1, sample(c(0, 1), private_students, replace =T)))

study_data %>%
  group_by(treatment) %>%
  summarize(sum(metro_card),
            n(),
            mean(metro_card))
```

Everyone without the treatment received a metro card and about half of the private school students received a free metro card through some other means.

#### While loops

Sometimes it is useful to do something until a certain condition is met. `while` loops can be used for this purpose. Suppose, for example, you wanted to sum all of the numbers from 0 to 100:

```{r}
i <- 0
total <- 0
while (i <= 100) {
  total <- total + i
  i <- i + 1
}
total
```
